---
title: "HW Class 12 Pt.2: Population Analysis"
author: "Ruth Barnes: A16747659"
format: pdf
---

# Class 12 Pt. 1: RNASeq Galaxy

Download and read a CSV from ENSEMBLE:

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

How many of each genotype are there?

```{r}
table(mxl$Genotype..forward.strand.)
```

Proportion or percent of total for each genotype:

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Let's now dig into this further.

# Class 12 Pt. 2: Population Analysis

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale.

**Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.**

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

***Q13. How many samples do we have?***

```{r}
nrow(expr)
```

**Sample size of genotypes and summary of expr:**

```{r}
table(expr$geno)
```

```{r}
summary(expr)
```

***Q13. Let's examine the three genotypes to find the median expression levels for each:***

```{r}
inden <- expr$geno == "G/G"
summary(expr[inden, "exp"])
```

```{r}
inden <- expr$geno == "A/A"
summary(expr[inden, "exp"])
```

```{r}
inden <- expr$geno == "A/G"
summary(expr[inden, "exp"])
```

***Q14: Generate a boxplot with a box per genotype***

Let's make a boxplot:

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(geno, exp) +
  geom_boxplot()
```

Let's make the boxplot nicer and easier to understand:

```{r}
ggplot(expr) + aes(geno, exp, fill = geno) +
  geom_boxplot(notch = T)
```
***Q14. What could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?***

The relative expression values between A/A and G/G show that the A/A genotype has the highest expression of ORMDL3, while G/G has the lowest expression of ORMDL3. This trend suggests that the SNP does affect the expression of ORMDL3, where the G allele is associated with lower expression levels.








