---
title: "Class 18: Pertussis Mini Project"
author: "Ruth Barnes: A16747659"
date: 02-06-2025
toc: true
format: pdf
---

# Pertussis and the CMI-PB project

## Background

Pertussis, (a.k.a) Whooping Cough is a deadly lung infection caused by the bacteria B. Pertussis. 

## 1. Investigating pertussis cases by year

The CDC tracks Pertussis cases around the US:
http://tinyurl.com/pertussiscdc

We can "scrape" this data using the R **datapasta** package:

***Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.***

```{r, include=FALSE}
cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L,2024),
  cases = c(107473,164191,165418,152003,
            202210,181411,161799,197371,
            166914,172559,215343,179135,265269,
            180518,147237,214652,227319,103188,
            183866,222202,191383,191890,109873,
            133792,109860,156517,74715,69479,
            120718,68687,45030,37129,60886,
            62786,31732,28295,32148,40005,
            14809,11468,17749,17135,13005,6799,
            7717,9718,4810,3285,4249,3036,
            3287,1759,2402,1738,1010,2177,2063,
            1623,1730,1248,1895,2463,2276,
            3589,4195,2823,3450,4157,4570,
            2719,4083,6586,4617,5137,7796,6564,
            7405,7298,7867,7580,9771,11647,
            25827,25616,15632,10454,13278,
            16858,27550,18719,48277,28639,32971,
            20762,17972,18975,15609,18617,
            6124,2116,3044,35493)
)
```

```{r}
head(cdc)
```

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(year, cases) +
  geom_line()
```

## 2. A tale of two vaccines (wP & aP)

Two types of pertussis vaccines have been developed: whole-cell pertussis (wP) and acellular pertussis (aP). The first vaccines were composed of ‘whole cell’ (wP) inactivated bacteria. The latter aP vaccines use purified antigens of the bacteria.

***Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine. What do you notice?***

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(year, cases) +
  geom_line() +
  geom_vline(xintercept = 1946, color = "blue", linetype = "dashed") + 
  geom_vline(xintercept = 1996, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2020, color = "darkgrey", linetype = "dashed") + 
  geom_vline(xintercept = 2003, color = "darkgreen", linetype = "dashed")
```
There were high case numbers before the first wP (whole-cell) vaccine roll out in 1946, then a rapid decline in case numbers until 2004, when we have our first large-scale outbreaks fo Pertussis again. There is also a notable COVID realted dip and recent rapid rise.

***Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?***

After the introduction of the aP vaccine, there is a peak in case numbers due to waning immunity and the aging of the vaccine (need booster shots). 

## 3. Computational Models of Immunity Pertussis Boost (CMI-PB)

### The CMI-PB API returns JSON data
 
The CMI-PB project aims to address this key question: What is different between the immune response to infection if you had an older wP vaccine vs the newer aP vaccine?

We can look at all the data from this ongoing project via JSON API calls. For this we will use the **jsonlite** package. We can install with: `install.packages("jsonlite")`

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject)
```
***Q. How many individials "subjects" are in this data set?***

```{r}
nrow(subject)
```

***Q4. How many aP and wP infancy vaccinated subjects are in the dataset?***

```{r}
table(subject$infancy_vac)
```

***Q5. How many Male and Female subjects/patients are in the dataset?***

```{r}
table(subject$biological_sex)
```

***Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?***

```{r}
table(subject$race, subject$biological_sex)
```

This is not representative of the US population but it is the biggest data-set of its type so let's see what we can learn...

### Side-Note: Working with dates

***Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?***

```{r}
library(lubridate)

# Calculate age in years using a fixed date
subject$age <- time_length(today("2025-03-08") - ymd(subject$year_of_birth), "years")

print(subject$age)
```

```{r}
library(dplyr)

# (i) Average age of aP individuals

wp <- subject %>% filter(infancy_vac == "wP")
avg_wp <- mean(wp$age, na.rm = TRUE)
print(avg_wp)
```

```{r}
# (ii) Average age of aP individuals
ap <- subject %>% filter(infancy_vac == "aP")
avg_ap <- mean(ap$age, na.rm = TRUE)
print(avg_ap)
```

```{r}
# (iii) Statistical significance test (t-test)
t_test <- t.test(wp$age, ap$age, var.equal = TRUE)

# Print results
cat("Average age of wP individuals:", avg_wp, "\n")
cat("Average age of aP individuals:", avg_ap, "\n")
print(t_test)
```

***Q8. Determine the age of all individuals at time of boost?***

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

***Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?***

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

### Joining multiple tables

Obtain more data from CMI-PB: 

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = T) 
ab_data <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = T)
```

```{r}
head(specimen)
```

```{r}
head(ab_data)
```

I now have 3 tables of data from CMI-PB: `subject`, `specimen`, and `ab_data`. I need to "join" these tables so I will have all the info u need to work with. 

For this we will use the `inner_join()` function from the **dplyr** package. 

***Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:***

```{r}
library(dplyr)

meta  <- inner_join(subject, specimen)
head(meta)
```

```{r}
dim(subject)
dim(specimen)
dim(meta)
```

Now we can join our `ab_data` table to `meta` so we have all the information we need about antibody levels.

***Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.***

```{r}
abdata <- inner_join(meta, ab_data)
head(abdata)
```

***Q11. How many different antibody isotypes are there in this dataset?/Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?***

```{r}
length(abdata$isotype)
```

```{r}
table(abdata$isotype)
```

```{r}
table(abdata$antigen)
```

***Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?***

```{r}
unique(abdata$dataset)
```
```{r}
table(abdata$dataset)
```

The most recent dataset, 2023_dataset, has 15,050 rows, which is an increase from 2021 and 2022 but still less than half of 2020’s total

I want a plot of antigen levels, across the whole dataset.

***Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:***

```{r}
library(ggplot2)

ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

Antigens like FIM2/3, PT, FELD1 have quite a large range of values. Others like Measles don't show much activity. 


***Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?***

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() + 
  facet_wrap(~infancy_vac)
```

Anitgens FIM2/3, LOS, and FELD1 show slightly similar levels of IgG antibody titers. These antigen levels are different due to variations in immune response, antigen exposure, and immune memory.

## 4. Examine IgG Ab titer levels

For this I need to select out just iostype IgG.

```{r}
igg <- abdata |>
  filter(isotype == "IgG")
head(igg)
```

A overveiw boxplot:
```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

Digging in further to look at the time course of IgG isotype PT antigen levels actoss aP and wP individauls:

```{r}
# Filter to include 2021 data only
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

# Fo;ter to look at IgG PT data only
abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  
# Plot and color by infancy_vac (wP vs aP)
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```










